---
title: "DS6372 Project 1"
author: "Caleb Thornsbury,  Stephanie Duarte, Steven Cox"
date: "2024-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# setwd("~/OneDrive - Southern Methodist University/Git_Website/DS6372_Project1")
library(ggthemes)
library(gridExtra)
library(ggplot2)
library(GGally)
library(dplyr)
library(tidyr)
library(knitr)
library(caret)
```

## Life Expectancy Data Set

```{r}
life <- read.csv("Data/Life_Expectancy_Data.csv")
summary(life)
sapply(life, class)

```

## Preliminary Data Wrangling

Explanatory variable is Life.expectancy. Therefore, before starting the EDA, any rows missing the explanatory variable will be filtered out.

```{r}
missing_life_expectancy <- life %>% select(Life.expectancy) %>% sapply(function(x) sum(is.na(x)))
print(missing_life_expectancy)
#Remove all rows that have missing values of our explanatory variable Life.expectancy
life <- life %>%
  filter(!is.na(Life.expectancy))

# To see the first few rows of the filtered data frame
head(life)
```

There are two categories that are not numerical, so lets take a look at the distributions of Countries and Status

```{r Distribution and uniqueness of Countries}
## Determine number of unique Countries
unique(life$Country)
## Create a table with number of observations
obs_per_country <- table(life$Country)  
## Convert to dataframe to visualize
df_obs_per_country <- as.data.frame(obs_per_country, stringsAsFactors = FALSE)
unique(df_obs_per_country$Freq)

```

-   All of the countries have the same number of observations, 16.

```{r Distribution of Status}
# Visualize the distribution of Status
ggplot(life, aes(x=Status)) +
  geom_bar(fill = "blue") +
  theme_minimal() +
  labs(title="Distribution of Status", x="Status", y="Count")
```

Section to describe the data set and explanatory variable.

## EDA
### Missing and NA Analysis
```{r}
sapply(life, function(x) sum(is.na(x)))
missing_count <- life %>% select(-Life.expectancy) %>% sapply(function(x) sum(is.na(x)))
missing_table <- as.data.frame(missing_count) %>%
  filter(missing_count >= 1) %>% arrange(desc(missing_count))
print(missing_table)

# Sort data frame by the number of missing values
# Adding an ID column to dataframe
life$ID <- 1:nrow(life)

# Calculate the number of missing values per observation
life$MissingCount = rowSums(is.na(life))

# Create a new table with ID and MissingCount
missing_values_summary <- data.frame(ID = life$ID, MissingCount = life$MissingCount)

# Sort the summary table by MissingCount in descending order
missing_values_summary <- missing_values_summary[order(-missing_values_summary$MissingCount), ]

# Display the sorted summary table
head(missing_values_summary)

```

Write something about the missing values.....

## Objective 1

Purpose: Display the ability to build regression models using the skills and discussions from Unit 1, 2, and 3 with the purpose of identifying key relationships and interpreting those relationships in an organized and clear fashion (Unit 4) .

```{r}

```

## Objective 2

Purpose: This objective is to go through a process to compare multiple models with the goal of developing a model that can predict the best and do well on future data. \*Use caret library to help ensure model comparisons are on a “apples to apples” level of comparison.  Consider setting the seed each time you run a new model and perform CV.

-   Feature Selection

```{r}

```

-   Cross-Validation

```{r}

```

-   Bootstrapping????

```{r}

```
