---
title: "DS6372 Project 1"
author: "Caleb Thornsbury,  Stephanie Duarte, Steven Cox"
date: "2024-02-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# setwd("~/OneDrive - Southern Methodist University/Git_Website/DS6372_Project1")
```

## Life Expectancy Data Set

```{r}
life <- read.csv("Data/Life_Expectancy_Data.csv")
```

Section to describe the data set and explanatory variable.

#### Preliminary EDA

```{r}

```

## Objective 1

Purpose: Display the ability to build regression models using the skills and discussions from Unit 1, 2, and 3 with the purpose of identifying key relationships and interpreting those relationships in an organized and clear fashion (Unit 4) .

```{r}

```

## Objective 2

Purpose: This objective is to go through a process to compare multiple models with the goal of developing a model that can predict the best and do well on future data. \*Use caret library to help ensure model comparisons are on a “apples to apples” level of comparison.  Consider setting the seed each time you run a new model and perform CV.

-   Feature Selection

```{r}

```

-   Cross-Validation

```{r}

```

-   Bootstrapping????

```{r}

```
