---
title: "DS6372 Project 1"
author: "Caleb Thornsbury, Stephanie Duarte, Steven Cox"
date: "2024-02-02"
output:
  html_document: 
    fig_caption: true
  # word_document: default
  # pdf_document:
  #   includes:
  #     in_header: header.tex  
  #   keep_tex: true
  #   latex_engine: xelatex
  #   geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
suppressPackageStartupMessages({
  library(knitr)
  library(kableExtra)
  library(ggthemes)
  library(gridExtra)
  library(ggplot2)
  library(GGally)
  library(dplyr)
  library(tidyr)
  library(caret)
  library(countrycode)
  library(car)
  library(olsrr)
  library(glmnet)
  library(naniar)
  library(lmboot)
})


```

```{r Clean Column Names, include=FALSE}
# Custom function to clean up any issues with column names
processColumnNames <- function(data_frame) {
  # Apply the custom name processing logic to each column name
  modifiedNames <- sapply(names(data_frame), function(name) {
    name <- tolower(name)  # Lowercase all characters
    name <- trimws(name)  # Strip leading and trailing spaces
    name <- gsub("_+", " ", name)  # Replace one or more underscores with a single space
    name <- gsub(" +", " ", name)  # Replace multiple spaces with a single space
    name <- gsub(" ", ".", name)  # Replace single space with a period
    name <- gsub("\\.{2,}", ".", name)  # Replace sequences of periods with a single period
    return(name)  # Return the modified name
  })
  # Ensure the modified column names are valid R identifiers and unique
  uniqueNames <- make.names(modifiedNames, unique = TRUE)
  names(data_frame) <- uniqueNames # Update the data frame's column names
  return(data_frame)
}
```

# Life Expectancy Data Set

This section explores the Life Expectancy Data Set. We aim to uncover insights into the factors affecting life expectancy across different predictors

```{r Import Dataset, warning=FALSE}
life_data <- read.csv(here::here("Data", "Life Expectancy Data.csv"))
# Clean up any column naming issues
life_data <- processColumnNames(life_data)
kable(head(life_data), caption = "A quick look at the life dataset")
```

# Objective 1

Purpose: Display the ability to build regression models using the skills and discussions from Unit 1, 2, and 3 with the purpose of identifying key relationships and interpreting those relationships in an organized and clear fashion (Unit 4) .

## Preliminary Data Wrangling

Explanatory variable is Life Expectancy. Therefore, before starting the EDA, any rows missing the explanatory variable will be filtered out.

```{r , warning=FALSE}
missing <- life_data %>% select(life.expectancy) %>% sapply(function(x) sum(is.na(x)))
print(paste("Number of observations missing explanatory variables:", missing))
#Remove all rows that have missing values of our explanatory variable life.expectancy
life_data <- life_data %>% filter(!is.na(life.expectancy))
```

```{r , warning=FALSE}

# Check uniqueness of each column
unique.predictors <- life_data %>% summarise(across(everything(), ~n_distinct(.)))
kable(unique.predictors, caption = "Unique variables per variable.")
# Take a look at the summary of the data set
# summary(life_data)
# dim(life_data)
```

### Missing and NA Analysis

```{r , warning=FALSE}
# Function to calculate percentage of NA values in a column
missing_percentages <- life_data %>% 
  summarise(across(everything(),~sum(is.na(.)))) %>%
  mutate_all(~ . / nrow(life_data) * 100) %>%
  rename_all(~ paste0(., "_percentage"))

sapply(life_data, function(x) sum(is.na(x)))

missing_count <- life_data %>% select(-life.expectancy) %>% sapply(function(x) sum(is.na(x)))

missing_table <- as.data.frame(missing_count) %>%
  filter(missing_count >= 1) %>% arrange(desc(missing_count))
print(missing_table)

# Create the missing data graph
vis_miss(life_data, sort_miss = TRUE)

```


This missing data report is crucial in the data cleaning and preprocessing phase of a data analysis pipeline. It helps to inform how to handle the missing values.

Checking life expectancy vs high missing data correlations

```{r , warning=FALSE}

# Collecting columns names that has missing data
col_missing_data <- row.names(missing_table)
# Checking correlation between the selected columns and life.expectancy
missing_cor <- cor(life_data[col_missing_data], 
  life_data[["life.expectancy"]], use = "complete.obs")
# Convert table to df so we can merge with the missing table
missing_cor <- data.frame(name =  rownames(missing_cor), Correlation = missing_cor)

missing_table$name <- rownames(missing_table)
missing_cor <- merge(missing_cor, missing_table, by = "name")
# We are only concerned with correlation, so taking absolute value
missing_cor$Correlation <- abs(missing_cor$Correlation)
# Arrange and display the table
missing_cor <- arrange(missing_cor, desc(Correlation))
kable(missing_cor, caption = "Missing Value Counts (Descending Order)", align = "c")

cor_threshold = 0.6 # Our determined threshold for what is correlated
filtered_corr <- missing_cor[missing_cor$Correlation < cor_threshold, ]
columns_to_remove <- filtered_corr$name
# Remove the columns from the data frame
life_filtered <- life_data[,!(names(life_data) %in% columns_to_remove)]
life_filtered$missing <- NULL
# revisit missing data with filtered data frame
sapply(life_filtered, function(x) sum(is.na(x)))
# Checking to see if all 160 remaining missing vars are the same observations
sum(is.na(life_filtered$income) & is.na(life_filtered$schooling))
# Removing additional missing data 
life_filtered <- life_filtered %>% na.omit()

```

```{r , warning=FALSE}

life_filtered %>%
  ggcorr(
    label = TRUE,
    label_size = 2,
    label_round = 2,
    hjust = 1,
    size = 3,
    color = "royalblue",
    layout.exp = 5,
    low = "darkorange",
    mid = "white",
    high = "darkorange",
    name = "Correlation") + 
  ggtitle("Inter-variable Correlation Matrix ") + theme_gdocs()

## Gather variable pairs that have a correlation greater than 0.8, as that could be an indication of multicollinearity. 
life_numeric <-  life_filtered %>%
  select(where(is.numeric)) %>% na.omit()
cor_matrix <- cor(life_numeric)
cor_pairs <- function(cor_matrix, threshold = 0.8) {
  cor_matrix[lower.tri(cor_matrix)] <- NA  # Mask lower triangle to avoid duplicate pairs
  # Use abs(cor_matrix) to consider the absolute value of correlations
  pairs <- which(abs(cor_matrix) > threshold & abs(cor_matrix) < 1, arr.ind = TRUE)
  pairs <- data.frame(Var1 = rownames(cor_matrix)[pairs[,1]],
                      Var2 = colnames(cor_matrix)[pairs[,2]],
                      Correlation = cor_matrix[pairs])
  # Ensure unique pairs by checking both combinations (Var1-Var2 and Var2-Var1)
  pairs <- pairs[!duplicated(paste(pmin(pairs$Var1, pairs$Var2), pmax(pairs$Var1, pairs$Var2))), ]
  return(pairs)
}
## View the significant pairs with high correlation
print(cor_pairs(cor_matrix))
life_filtered$infant.deaths <- NULL

```

### Distribution of Status

```{r Distribution of Status, warning=FALSE}

# Calculate counts and percentages
summary_df <- life_filtered%>%
  group_by(status) %>%
  summarise(count = n()) %>%
  mutate(percentage = round(count / sum(count) * 100, 2),
         label = paste0(percentage, "%")) %>%
  ungroup()

# Visualize the distribution of Status with percentages
ggplot(life_filtered, aes(x = status)) +
  geom_bar(fill = "peachpuff", aes(y = after_stat(count))) + 
  geom_text(data = summary_df, aes(x = status, y = count/2, label = label), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  theme_clean() +
  labs(title = "Distribution of Status", x = "Status", y = "Number of Observations")


ggplot(life_filtered, aes(x = status, y = life.expectancy)) +
  geom_boxplot(fill = "peachpuff") +
  labs(x = "Status", y = "Life Expectancy") +
  theme_minimal() +
  coord_flip()  # Rotate the plot


```

Developed 17% , Developing 83%.

### Visualize Life Expectancy over the Years

```{r Plot Life vs Years, warning=FALSE}
ggplot(life_filtered, aes(x = year, y = life.expectancy)) +
  geom_smooth() +
  theme_classic() +
  labs(title = "Life Expectancy Over Years",
       x = "Year",
       y = "Life Expectancy")
```

### Number of observations per Country

```{r Distribution and uniqueness of Countries, results='hide', warning=FALSE}
## Create a table with number of observations
table(life_filtered$country)
```

-   All of the countries have the same number of observations, 16. \### Visualize Life Expectancy over the Years by Region

```{r Plot LifeVsYears by Region, warning=FALSE}
# Use the country code package to break up the Countries by Region
life_filtered$region <- as.factor(countrycode(life_filtered$country, "country.name", "region"))

ggplot(life_filtered, aes(x = year, y = life.expectancy, group = region, color = region)) +
  geom_smooth() +
  theme_classic() +
  labs(title = "Life Expectancy Over Years by Region",
       x = "Year",
       y = "Life Expectancy") +
  theme(legend.position = "bottom")

```

There is definitely a trend to consider when considering life expectancy based on which region the person is from. Might consider using a dummy variable to break up region numerically for modeling.

### Checking correlation and variance.

##Visualization on response variable and correlated predictors

```{r , warning=FALSE}

ggplot(data = life_filtered, aes(x = life.expectancy)) + 
  geom_histogram(binwidth = 1, fill = "lightblue", color = "black") + 
  labs(title = "Histogram of Life Expectancy", x = "Life Expectancy", y = "count") +
  theme_minimal()

ggplot(data = life_filtered, aes(y = life.expectancy, x = schooling)) +
  geom_point(color = "blue") +
  geom_smooth(color="red")+
  theme_clean()+
  labs(title = "Distribution of Schooling",
       y = "Life Expectancy",
       x = "Number of Years of Schooling")

ggplot(data = life_filtered, aes(y = life.expectancy, x = adult.mortality)) +
  geom_point(color = "coral") +
  geom_smooth(color="red")+
  theme_clean()+
  labs(title = "Distribution of Adult Mortality Rates",
       y = "Life Expectancy",
       x = "Adult Mortality")

ggplot(data = life_filtered, aes(y = life.expectancy, x = income.composition.of.resources)) +
  geom_point(color = "darkcyan") +
  geom_smooth(color="red")+
  theme_clean()+
  labs(title = "Distribution of Income Composition of Resources",
       y = "Life Expectancy",
       x = "Income Composition of Resources")

ggplot(data = life_filtered, aes(y = life.expectancy, x = hiv.aids)) +
  geom_point(color = "lightgreen") +
  geom_smooth(color="red")+
  theme_clean()+
  labs(title = "Distribution of HIV/AIDS",
       y = "Life Expectancy",
       x = "Income Composition of Resources")
```

## Preparing data set for modeling

```{r, warning=FALSE}
life_filtered$country <- NULL
life_filtered$status <- as.factor(life_filtered$status)

```

```{r, initial_investigative_model, echo=FALSE, warning=FALSE}
#First model to find important variables
simple.fit <-lm(life.expectancy ~ ., data = life_filtered)
summary(simple.fit)
vif(simple.fit)

simple.fit.anova <- aov(simple.fit)
summary(simple.fit.anova)

# par(mfrow=c(2,2)) # Set up a 2x2 plot grid
# plot(simple.fit, which = 1) # Residuals vs Fitted
# plot(simple.fit, which = 2) # Normal Q-Q
# plot(simple.fit, which = 4) # Cook's distance
# par(mfrow = c(1, 1)) # Set up a 1x1 plot grid

ols_plot_diagnostics(simple.fit)

# Extracting coefficients and their p-values----
coefficients_summary <- summary(simple.fit)$coefficients
# Define significant predictors
significant_predictors <- c("region", "year", "status", "adult.mortality", "percentage.expenditure",
                            "under.five.deaths", "hiv.aids", "schooling", "income.composition.of.resources")
# Dynamically create the formula
new_formula_str <- paste("life.expectancy ~", paste(significant_predictors, collapse = " + "))
# Convert string to formula
new_formula <- as.formula(new_formula_str)
# new_formula is now a formula object that can be used in modeling functions
print(as.formula(new_formula_str))

# Refitting the model
simple.fit2<-lm(as.formula(new_formula_str),data=life_filtered)
summary (simple.fit2)
print(simple.fit2)


simple.fit.anova2 <- aov(simple.fit2)
summary(simple.fit.anova2)

ols_plot_diagnostics(simple.fit2)

#Check the residuals

# plot(simple.fit2, which = 1) # Residuals vs Fitted
# plot(simple.fit2, which = 2) # Normal Q-Q
# plot(simple.fit2, which = 4) # Cook's distance
# cutoff <- 4/(nrow(life_filtered)-length(simple.fit2$coefficients)-2)
# plot(simple.fit2, which=4, cook.levels=cutoff)
# abline(h=cutoff, lty=2, col="red")

```

## Final Basic Linear Regression Model

### Split the Dataset

Splitting up the data into a training and testing set so that we can compare the different models using the same data.

```{r, warning=FALSE}
#training and test set 80/20 split

set.seed(1234)
trainIndex <- createDataPartition(life_numeric$life.expectancy, p = .80, list = FALSE)
train <- life_filtered[trainIndex, ]
test <- life_filtered[-trainIndex, ]

```

```{r , warning=FALSE}

model_train <- lm(new_formula, data = train)

# Predictions and Evaluation

predictions <- predict(model_train, newdata = test)

rmse_value <- sqrt(mean((predictions - test$life.expectancy)^2))

print(paste("Basic Model RMSE:",rmse_value))

```

RMSE for the test set: 3.81

# Objective 2

Purpose: This objective is to go through a process to compare multiple models with the goal of developing a model that can predict the best and do well on future data. \*Use caret library to help ensure model comparisons are on a “apples to apples” level of comparison.

### Feature Selection

```{r, warning=FALSE}

set.seed(1234)
fitControl<-trainControl(method="repeatedcv",number=10,repeats=1) 
glmnet.fit<-train(new_formula,
               data=life_filtered,
               method="glmnet",
               trControl=fitControl
               )
opt.pen<-glmnet.fit$finalModel$lambdaOpt 
coef(glmnet.fit$finalModel,opt.pen)
glmnet.fit

set.seed(1234)
trainIndex <- createDataPartition(life_filtered$life.expectancy, p = .80, list = FALSE)
train <- life_filtered[trainIndex, ]
test <- life_filtered[-trainIndex, ]

model_train <- lm(new_formula, data = train)

summary(model_train)

# # Plot residuals
# par(mfrow=c(2, 2)) # Set up the plotting area to display multiple plots
# plot(model_train)

# Predictions and Evaluation
predictions <- predict(model_train, newdata = test)
rmse_value <- sqrt(mean((predictions - test$life.expectancy)^2))
print(paste("Feature Model RMSE:",rmse_value))

```

Feature Model RMSE: 3.78

### 10-fold cross validation to determine an appropriate $k$

```{r, warning=FALSE}

set.seed(1234)  # Set the seed for reproducibility

# Create a grid to investigate the values of k
k_values <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 20, 30)
tuneGrid <- expand.grid(k = k_values)

# Set up 10-fold cross-validation
fitControl <- trainControl(method = "cv",
                           number = 10,
                           summaryFunction = defaultSummary)

# Fit the kNN model
knn_fit <- train(new_formula, 
                 data = life_filtered, 
                 method = "knn", 
                 tuneGrid = tuneGrid, 
                 trControl = fitControl,
                 preProcess = c("center", "scale","YeoJohnson", "nzv", "pca"))  # Standardize predictors

# Output the results
print(knn_fit)

# # Plot residuals
# par(mfrow=c(2, 2)) # Set up the plotting area to display multiple plots
# plot(knn_fit)
```

KNN Model RMSE 2.27

# Table of RMSE results
```{r, results='asis'}
model_data <- data.frame( Model = c("MLN", "Feature Selection (Glmnet)", "KNN 10 Fold"), RMSE = c(3.81, 3.78, 2.27) )
kable(model_data, caption = "Model Performance Comparison", align = c('l', 'r'))
```

BOOTSTRAPPING

```{r, warning=FALSE}

#residual bootstrapping
boot.res<-residual.boot(new_formula,data=life_filtered,B=10000,seed=1234)
t(apply(boot.res$bootEstParam,2,quantile,probs=c(.025,.975)))

# #paired bootstrapping
boot.p<-paired.boot(new_formula,data=life_filtered,B=10000,seed=1234)
t(apply(boot.p$bootEstParam,2,quantile,probs=c(.025,.975)))

```

# Appendix A

```{r Appendix_A, warning=FALSE}

# ggplot(life_filtered, aes(x = year, y = life.expectancy, group = country, color = country)) +
#   geom_smooth() +
#   theme_classic() +
#   labs(title = "Life Expectancy Over Years by Country",
#        x = "Year",
#        y = "Life Expectancy") +
#   theme(legend.position = "none")
```
